The finite representation property for some reducts of relation algebras

### 2020-12-22 frp-paper/Paper.tex
the R(;,+) case (wip): fixes and thoughts

@@

Let $\mathcal{A}$ is a join-semilattice ordered semigroup, a subset $F \subseteq \mathcal{A}$ is called \emph{prime filter}, if $F$ is upwardly closed and $x + y \in F$ implies $x \in F$ or $y \in F$. The spectrum of $\mathcal{A}$, denoted as $\operatorname{Spec}(\mathcal{A})$, is the set of all prime filters.

\begin{definition} Let $\mathcal{A}$ be a join-semilattice ordered semigroup. A prenetwork over $\mathcal{A}$ is a pair $(E, V, N)$, where $E$ is a set of nodes, $V$ is a set of nodes (an equivalence relation on $E$), and $N$ is a function $l : V \to \operatorname{Spec}(\mathcal{A})$.

  A prenetwork over $\mathcal{A}$ is a network if the following hold:
  \begin{enumerate}
    \item For every $u, v \in E$ and for every $x,y,z \in \mathcal{A}$ $z \in l(u, v)$ and $z \leq x \: ; \: y$ implies $x \in N(u, w)$ and $y \in l(w, v)$ for some $w \in E$,
    \item For every $u, v, w \in E$ one has $l(u, v) ; l(v, w) \subseteq l(u, w)$,
    \item For every $u, v \in E$ and for every $x,y,z \in \mathcal{A}$ $z \in l(u, v)$ and $z \leq x + y$ implies either $x \in l(u, v)$ or $y \in l(u, v)$.
  \end{enumerate}
\end{definition}

Let $\mathcal{N}_1 = (E_1, V_1, l_1)$ and $\mathcal{N}_2 = (E_2, V_2, l_2)$ be (pre)networks, then $\mathcal{N}_1$ is a \emph{subnetwork} of $\mathcal{N}_2$, written as $\mathcal{N}_1 \subseteq \mathcal{N}_2$,
if $E_1 \subseteq E_2$, $V_1 \subseteq V_2$, and for all $x, y \in E_1$ if $a \in l_1(x, y)$ then there is $b \in l_2(x, y)$ with $a \leq b$.

@@

  \begin{enumerate}
    \item Round $0$: $\forall$ picks a prime filter $F_0$ of $(\mathcal{A})$. $\exists$ responds with a prenetwork $\mathcal{N}_0 = (E_0 = \{ x_0, x_1 \}, V_0, l_0)$ such that the $l_0(x_0, x_1) = l_0(x_1, x_0) = F_0$ and $l_0(x_0, x_0) = l_0(x_1, x_1) = {\uparrow {\bf 1}}$.
    \item Round $n + 1$. Suppose, the prenetwork $\mathcal{N}_n = (E_n, V_n, l_n) \rangle$ has been played and $F_n \in \operatorname{Spec}(\mathcal{A})$.

    $\forall$ picks elements $a, b, c \in \mathcal{A}$ with $a \: ; \: b \geq c$

    $\exists$ responds with the network $\mathcal{N}_{n + 1} = (E_{n + 1}, V_{n + 1}, l_{n + 1})$
  \end{enumerate}

@@

  \begin{enumerate}
    \item If $\mathcal{A}$ is representable then $\exists$ have a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$.
    \item If $|\mathcal{A}| = \omega$ and $\exists$ have a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is representable.
    \item For each $n < \omega$ there exists a first-order sentence $\sigma_n$ such that $\exists$ have a winning strategy in $\mathcal{G}_n(\mathcal{A})$ iff $\mathcal{A} \models \sigma_n$.
  \end{enumerate}
\end{lemma}

@@

  \begin{enumerate}
    \item Let $h$ be a representation of $\mathcal{A}$
    \item
    \item
  \end{enumerate}
\end{proof}

@@

\begin{proof}
  Let $\mathcal{A} \in {\bf R}(+, ;)$ such that $|\mathcal{A}| < \omega$ and $h$ a representation of $\mathcal{A}$
\end{proof}

\section{Acknowledgements}

The author is sincerely grateful to Robin Hirsch, Ian Hodkinson, Stepan Kuznetsov, Valentin Shehtman, Eugeny Zolin, and his supervisor Ilya Shapirovsky for valuable conversations.\\The previous version contained an error. Thanks to Jas Semrl for noticing that.

>>>>

Let $\langle P, \leq \rangle$ be a partial order. An upper cone generated by $x$ is the set $\uparrow x = \{ a \in P \: | \: x \leq a \}$. Let $A \subseteq P$, then $\uparrow A = \bigcup \limits_{x \in A} \uparrow x = \{ a \in P \: | \: \exists x \in P \: x \leq a \}$. The set of all upper cones of a poset $\langle P, \leq \rangle$ is denoted as $\operatorname{Up}(P)$.

\subsection{Axiomatisating ${\bf R}(+, ;)$ and the finite representation property}

@@

\begin{definition} Let $\mathcal{A}$ be a join-semilattice ordered semigroup. A prenetwork over $\mathcal{A}$ is a pair $(E, V, N)$, where $E$ is a set of nodes, $V$ is a set of nodes such that $\langle E, V \rangle$ is a complete directed graph, and $N$ is a function $l : V \to \operatorname{Up}(\mathcal{A})$.

  A prenetwork over $\mathcal{A}$ is a network if the following hold:
  \begin{enumerate}
    \item For every $u, v \in E$ and for every $x,y,z \in \mathcal{A}$ $z \in l(u, v)$ and $z \leq x \: ; \: y$ implies $x \in N(u, w)$ and $y \in l(w, v)$ for some $w \in E$.
    \item For every $u, v, w \in E$ one has $l(u, v) ; l(v, w) \subseteq l(u, w)$.
    \item For every $u, v \in E$ $l(u,v)$ is join-prime. That is, for each $a, b \in \mathcal{A}$ if $a + b \in l(u,v)$, then either $a \in l(u,v)$ or $b \in l(u,v)$.
  \end{enumerate}
\end{definition}

Let $\mathcal{N}_1 = (E_1, V_1, l_1)$ and $\mathcal{N}_2 = (E_2, V_2, l_2)$ be (pre)networks, then $\mathcal{N}_1$ is a \emph{subnetwork} of $\mathcal{N}_2$, written as $\mathcal{N}_1 \subseteq \mathcal{N}_2$,
if $E_1 \subseteq E_2$, $V_1 \subseteq V_2$, and for all $x, y \in E_1$ if $a \in l_1(x, y)$ then there is $b \in l_2(x, y)$ with $b \leq a$.

@@

  \begin{enumerate}
    \item Round $0$: $\forall$ picks elements $a, b \in \mathcal{A}$ with $a \not\leq b$. $\exists$ responds with a prenetwork $\mathcal{N}_0 = (E_0 = \{ x_0, x_1 \}, V_0 = \{ (x_0, x_1)\}, l_0)$ such that $l_0(x_0, x_1) = \uparrow b$.
    \item Round $n + 1$. Suppose, the prenetwork $\mathcal{N}_n = (E_n, V_n, l_n)$ has been played.

    $\forall$ has the following three options:
    \begin{enumerate}
      \item ({\bf Composition move}): $\forall$ can pick $x, y, z \in E_n$ with $a \in l_n(x, y)$ and $b \in l_n(y, z)$. We denote such a move as $N(x,y,z,a,b)$. Then $\exists$ responds with $\mathcal{N}_{n + 1} = (E_{n + 1}, V_{n + 1}, l_{n + 1})$ such that $\mathcal{N}_{n + 1}$
      is the same as $\mathcal{N}_n$, but $l_{n + 1}(x, z) = \uparrow (l_{n}(x, z) \cup \{ a \: ; \: b\})$.
      \item ({\bf Witness move}):

      $\forall$ picks nodes $x, y \in E_n$ and $a, b \in \mathcal{A}$ such that $c \leq a ; b$ for $c \in l_n(x, y)$. $\exists$ has to find a witness. She responds with a fresh node $z$. We define the prenetwork $T$ , the edges of which are $x, y, z$ with the following labelling:
      \begin{enumerate}
        \item $l_T(x, z) = \uparrow a$
        \item $l_T(z, y) = \uparrow b$
      \end{enumerate}
      For her response, $\exists$ plays $N_{n + 1} = N_n \cup T$
      \item ({\bf Join move}):

      $\forall$ picks nodes $x, y \in E_n$ and $a + b$ for $a, b \in \mathcal{A}$. $\exists$ has the following two alternatives for her response:
      \begin{enumerate}
        \item $\exists$ chooses $a$ and responds with the prenetwork $\mathcal{N}_{n + 1} = \langle E_n, V_{n}, l_{n + 1} \rangle$, where $l_{n+1}(x, y) = \uparrow (l_n(x,y) \cup \{ a \})$.
        \item $\exists$ chooses $b$. The response is similar but $l_{n + 1}(x, y) = \uparrow (l_n(x,y) \cup \{ b \}$.
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}

@@

  \begin{enumerate}
    \item If $\mathcal{A}$ is representable then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$.
    \item If $|\mathcal{A}| = \omega$ and $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is representable.
  \end{enumerate}
\end{lemma}

@@

  \begin{enumerate}
    \item Let $h$ be a representation of $\mathcal{A}$ over some domain $D$. $\forall$ plays $(a, b)$ with $a \not\leq b$ for the initial round. Let $N = \langle E, V, l \rangle$ be a network being played.

    TODO: define an embedding $\iota : E \to D$ and check the presence of a winning strategy explicitly

    \item Let $a, b \in \mathcal{A}$ such that $a \not\leq b$. We consider a play of the game where $\forall$ pick the pair $(a, b)$ in the initial round and plays $(\mathcal{N},x, y, z, c, d)$ in the further rounds for each $x, y, z \in \operatorname{Nodes}(\mathcal{N})$ and $c,d \in \mathcal{A}$. Here, $c \in l_N(x, y)$ and $d \in l_N(y, z)$.

    $\forall$ also plays all rounds $(\mathcal{N},x,y,c,d)$ for each $x, y \in \operatorname{Nodes}(\mathcal{N})$ and $c,d \in \mathcal{A}$ such that there is $e \in \mathcal{A}$ such that
    $e \leq c ; d$ and $e \in l_N(x,y)$.

    $\forall$ picks also $c + d$ and vertices $x,y \in \operatorname{Nodes}(\mathcal{N})$ for $c, d \in \mathcal{A}$.

    We have the following play of a game that we define as a sequence of prenetworks where $\exists$ has a winning strategy for each $i < \omega$:

    \begin{center}
      $\mathcal{N}_0 \subseteq \mathcal{N}_1 \subseteq \mathcal{N}_2 \dots$
    \end{center}

    Let us put $\mathcal{N} = \bigcup \limits_{i < \omega} \mathcal{N}_i$.

    Clearly that every $l_{\mathcal{N}}(x, y)$ for each $x, y \in \operatorname{Nodes}(\mathcal{N})$ is join-prime. Let $a + b \in l_{\mathcal{N}}(x, y)$, then there exists $i < \omega$ with
    $a + b \in l_{\mathcal{N}_i}(x, y)$

    $\forall$ makes all possible join moves, so either $a \in l_{\mathcal{N}_{i + 1}}(x, y)$ or $b \in l_{\mathcal{N}_{i + 1}}(x, y)$ depending on what $\exists$ chooses.

    Let us ensure that the labelling function $l_{\mathcal{N}}$ respects $;$. Here, the argument is similar to the proof of \cite[Proposition 5]{hirsch2005class}.

    The map $rep$ defined as follows is a representation:
    \begin{center}
      $rep(a) = \{ (x, y) \: | \: a \in l_{\mathcal{N}}(x, y)\}$
    \end{center}
    This map is well-defined.

    $rep(a + b) = \{ (x, y) \: | \: a + b \in l_{\mathcal{N}}(x, y)\} \subseteq \{ (x, y) \: | \: a \in l_{\mathcal{N}}(x, y) \lor b \in l_{\mathcal{N}}(x, y) \} = rep(a) \cup rep(b)$.

    The converse inclusion:

    Suppose $(x,y) \in rep(a) \cup rep(b)$, so either $a \in l_{\mathcal{N}}(x,y)$ or $b \in l_{\mathcal{N}}(x,y)$. If $a \in l_{\mathcal{N}}(x,y)$, then $a + b \in l_{\mathcal{N}}(x,y)$ since $a \leq a + b$ and $l_{\mathcal{N}}(x,y)$ is an upset. The last holds because
    $l_{\mathcal{N}}(x,y) = \bigcup \limits_{i < \omega} l_{\mathcal{N}_i}(x,y)$ and the union of upsets is an upset.

    The second alternative ($b \in l_{\mathcal{N}}(x,y)$) is similar to the second one.

    Thus, $rep(a + b) = rep(a) \cup rep(b)$.

\medskip

Let us check that $rep(a ; b) = rep(a) | rep(b)$.

$rep(a;b) = \{ (x,y) \: | \: a;b \in l_{\mathcal{N}}(x,y) \}$

Suppose $(x,y) \in rep(a) ; rep(b)$, we need $(x, y) \in rep(a ; b)$, i.e. $a;b \in l_{\mathcal{N}}(x,y)$. Thus, there exists $z \in \operatorname{Nodes}(\mathcal{N})$ such that $(x,z) \in rep(a)$ and $(z,y) \in rep(b)$, i.e. $a \in l_{\mathcal{N}}(x,z)$ and $b \in l_{\mathcal{N}}(z, y)$. $\mathcal{N}$ is a network, so $l_{\mathcal{N}}(x,z) ; l_{\mathcal{N}}(z,y) \subseteq l_{\mathcal{N}}(x,y)$. Thus, $a ; b \in l_{\mathcal{N}}(x,y)$.

Suppose for the converse, $(x, y) \in rep(a ; b)$, i.e. $a;b \in l_{\mathcal{N}}(x,y)$. We need some $z$ such that $(x, z) \in rep(a)$ and $(z, y) \in rep(b)$, i.e. $a \in l_{\mathcal{N}}(x,z)$ and $b \in l_{\mathcal{N}}(z,y)$. (Ideally, we need some $c$ with $c \leq a ; b$, but $l_{\mathcal{N}}(x,y)$ is not downward directed)
  \end{enumerate}
\end{proof}

\begin{definition}
  Let $V = \{ v_0, v_1, ... \}$ be a countable set of variables. The set of terms $T$ is generated by the following grammar:

  \begin{center}
    $T ::= V \: | \: (T + T) \: | \: (T ; T)$
  \end{center}
\end{definition}

\begin{definition}
  A term network is a graph $\langle E, V, l \rangle$, where $\langle E, V \rangle$ is a complete directed graph and $l : V \to 2^T$ is a labelling function.
\end{definition}

Let $\mathcal{A}$ be a join semilattice-ordered semigroup and $\vartheta : V \to \underline{{\mathcal{A}}}$ a valuation. The value of complex terms is defined inductively for $a, b \in T$:

\begin{enumerate}
  \item $(a ; b)^{\vartheta} = a^{\vartheta} ; b^{\vartheta}$
  \item $(a + b)^{\vartheta} = a^{\vartheta} + b^{\vartheta}$
\end{enumerate}

Let $\mathcal{N} = \langle E, V, l \rangle$ be a term prenetwork, $\mathcal{A}$ be a join-semilattice ordered semigroup and $\vartheta : V \to \underline{{\mathcal{A}}}$ a valuation. Let us define the prenetwork $\mathcal{N}^{\vartheta}$ with the same edges and vertices with the labelling $l^{\vartheta}(x, y) = \{ t^{\vartheta} \: | \: t \in l(x, y)\}$.

\begin{lemma}
  For each $n < \omega$ there exists a first-order sentence $\sigma_n$ such that $\exists$ has a winning strategy in $\mathcal{G}_n(\mathcal{A})$ iff $\mathcal{A} \models \sigma_n$.
\end{lemma}

\begin{proof}

\end{proof}

@@

\begin{proof}
  Let $\mathcal{A} \in {\bf R}(+, ;)$ such that $|\mathcal{A}| < \omega$ and $h$ a representation of $\mathcal{A}$. monkahmm
\end{proof}

\begin{comment}
\section{Acknowledgements}

The author is sincerely grateful to Robin Hirsch, Ian Hodkinson, Stepan Kuznetsov, Valentin Shehtman, Eugeny Zolin, and his supervisor Ilya Shapirovsky for valuable conversations.\\The previous version of this paper contained an error. Many thanks to Jas Semrl for noticing that.
\end{comment}
